---
title: "Figures_Ch.4"
author: "Cynthia Barile"
output:
  html_document:
    theme: journal
    highlight: breezedark
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: 
      collapsed: false # to prevent collapsing down smaller headers
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE)

## The lines below are to modify the style of the document, in particular here to extend the margins in the HTML
```

<style type="text/css">
.main-container {
  max-width: 1800px !important;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r packages}
# Set wd to the folder containing this script
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

library(dplyr)
library(embarcadero)
library(foreign)
library(giscoR) # to access GISCO data directly from R
library(ggplot2)
library(ggspatial)
library(grid)
library(marmap) # To open bathymetry data
library(sf)
library(raster)
library(RColorBrewer)
library(terra)
library(tmap)
```

Figures for "Chapter 3: Modelling SPW habitat use using PAM data" are generated in this script. Only Fig. 2 is created in PowerPoint and won't be found in here. 

# Fig. 1: Survey tracklines
Here we make a map for the PhD thesis (and paper), for the sperm whale PAM modelling study. 
First, start with the bathymetry, which is the base layer of the map:

```{r PAM bathymetry, results='hide', fig.keep = 'all'}
# BATHYMETRY RASTER
bath <- getNOAA.bathy(lon1 = -3, lon2 = -23, 
                      lat1 = 47, lat2 =58, resolution=1) # res in minutes
projection <- "+proj=utm +zone=29 +datum=WGS84"

# Study area box
area.box <- as(extent(-19, -5, 48, 57), 'SpatialPolygons') # create a box as a Spatial Object
crs(area.box) <- "+proj=longlat +datum=WGS84 +no_defs" # define the crs of this box
box <- spTransform(area.box, CRS("+init=epsg:32629")) # project the spatial object (WGS84 UTM Z29: EPSG 32629)

# Project the bathymetry raster
r1 <- marmap::as.raster(bath) # convert to a raster layer
r2 <- projectRaster(r1, crs=projection) # project it

# Crop the raster r2, using the box previously created
depth.proj <- crop(r2, box)
depth.proj[depth.proj > 0] <- NA # Remove the values for the land

contour.lines <- rasterToContour(depth.proj) # Get bathymetry lines

# create palette
greys <- rev(colorRampPalette(RColorBrewer::brewer.pal(9,"Greys"))(100))

tmap_mode("plot")

p <- tm_shape(depth.proj) +
  tm_raster(style='cont', palette=greys, 
            colorNA = 'grey80', alpha=.9, title = '', legend.show = FALSE) +
  tm_shape(contour.lines) + # Plot the contour lines
    tm_lines(lwd = 0.3, col='grey80', alpha=.6)
print(p)
```

Let's add the boxes delimiting the survey regions on top:

```{r PAM boxes, results='hide', fig.keep = 'all'}
# BOXES
northbox <- st_read("F:/PhD/ObSERVE_MASTER/ObSERVE_PAM/Shapefiles/NorthBox.shp")
southbox <- st_read("F:/PhD/ObSERVE_MASTER/ObSERVE_PAM/Shapefiles/SouthBox.shp")

# Project to the same crs as for the rest
northbox <- st_transform(northbox, projection)
southbox <- st_transform(southbox, projection)

p <- 
  p + tm_shape(northbox) +
  tm_fill(alpha=.5) +
    tm_borders(col="grey") +
  tm_shape(southbox) +
    tm_fill(alpha=.5) +
    tm_borders(col="grey")

print(p)
```

The next step is to add the transect lines:

```{r PAM transects, results='hide', fig.keep = 'all'}
# TRANSECT DATA
transects <- read_sf("../01_Data/PAM_Surveys.shp") # transect lines
sf::st_crs(transects) <- 4326 # set coordinate system (espg code of WGS84)
transects <- st_transform(transects, projection) # project
transects$Id <- c(1:nrow(transects)) # unique Id for each line (because some lines are split)

transects$Survey <- as.factor(transects$Survey) # factors to color by survey
palette.PAM <- c("#557A95", "#E2B4BD", "#9DD9D2", "#ECC783", "#956D90", "#91C0DC") # palette for the transect lines

p <- p +
  tm_shape(transects) +
    tm_lines(col="Survey", palette = palette.PAM, lwd = 2, legend.col.show = F)

print(p)
```

Now the sperm whale detections:
```{r detections}
det <- read.dbf("../01_Data/SPW_detections.dbf")
det <- det[-17,] # remove that sighting that falls outside of the survey box
det <- st_as_sf(det, coords=c("TMLong", "TMLat")) # transform as spatial object
st_crs(det) <- 4326 # set the crs
det <- st_transform(det, projection) # project

p <- p + 
  tm_shape(det) +
    tm_symbols(size = 0.1, col = "black", border.col = "black", alpha = 0.4, 
               border.lwd = 0.1, legend.col.show = F)

print(p)
```

The contents of the map are all there. We can now add necessary elements, such as the north arrow, scale, and modify the way the legend is displayed.

```{r map elements, results='hide', fig.keep = 'all'}
p1 <- p +
  tm_graticules(lines = FALSE) +
  tm_compass(position=c("left", "top"), size=2) +
  tm_scale_bar(position=c("RIGHT", "BOTTOM"), breaks = c(0,100,200), lwd = 0.1)

print(p1)
```

We create the legend in another object because I want to put them altogether in PhotoShop, it's too much of a hassle to adjust positions in the viewport.

First the legend for the bathymetry needs to be retrieved: 

```{r legend}
# Need to get the depth legend via the tm_raster and not with tm_add_legend
# otherwise it does not put a continuous gradient in legend but categories
legend.depth <- 
  tm_shape(depth.proj) +
  tm_raster(style='cont', palette=greys, 
# set showNA = F or it adds something for missing data in the legend
            colorNA = 'grey80', alpha=.9, title = "Depth (m)", showNA = FALSE) +
  tm_shape(contour.lines) + # Plot the contour lines
    tm_lines(lwd = 0.3, col='grey80', alpha=.6) +
  tm_layout(legend.only = TRUE, legend.text.fontfamily = "serif", 
            legend.title.fontface = "bold.italic", legend.title.fontfamily = "serif",
            legend.bg.color = "white", legend.text.size = .7, legend.title.size = 1) 
# legend.only, to retrieve the legend only

print(legend.depth)
```

And the legend created manually for the rest:

```{r }
legend <-  p +
  tm_add_legend(type="fill", labels = "  Survey regions", col = "grey", 
                alpha =0.5, border.col="grey") +
  tm_add_legend(type = "symbol", size = 1.03, col = "black", alpha = 0.4,
                border.col = "black", border.lwd = 0.2, 
                labels = "  Sperm whale detection events") +
  tm_add_legend(type="line", col = palette.PAM, 
                labels = c("  May 2015", "  Aug. 2015", "  Sep. 2015", "  May 2016", 
                           "  Jun. 2016", "  Oct. 2016"), lwd = 2, title = "Transect lines") +
  tm_layout(legend.title.fontface ="bold.italic", legend.title.size = 1.2, 
            legend.text.size = .9, legend.bg.color = "white", 
            legend.title.fontfamily = "serif", legend.only = TRUE, 
            legend.text.fontfamily = "serif")

print(legend)
```

Let's prepare an inset to add to the map. To do that, a shapefile of Europe is downloaded from GISCO and a box representing the study area is created below.

```{r}
# get the countries of europe - RG for having polygons
europe <- gisco_get_countries(region = "Europe", year = "2020", epsg = "4326",
                              resolution = 10, spatialtype = "RG")
europe <- st_transform(europe, projection) # project it
europe <- st_crop(europe, ymin=3444428, ymax=8027428, xmin=-1425598, xmax=4001958)

surveyarea.bb <- st_as_sfc(st_bbox(depth.proj)) # create bounding box

insetmap <- tm_shape(europe) +
  tm_borders(lwd = 0.5, col = "black") +
  tm_fill(col="grey") +
tm_shape(surveyarea.bb) +
  tm_borders(col = "red", lwd = 2) +
  tm_layout(inner.margins = c(0,0,0,0), outer.margins=c(0,0,0,0))

print(insetmap)
```

Let's save all the elements to work on the final product in Photoshop.

```{r export, eval = FALSE}
# print depth.proj to do xmax - xmin / ymax - ymin
asp <- (798115.9 - (-246004.1))/(6361188 - 5323338)

tmap_save(p1, filename="../03_Outputs/PAM_map.svg",
          dpi=900, height=asp*15, width=15, units="cm")
tmap_save(legend.depth, filename="../03_Outputs/legend_depth.svg",
          dpi=400, scale = 1, width=5, units="cm")
# `legend` I actually saved directly exporting from the plot window,
# using tmap_save it was fucking up the size of the symbol for the detections
tmap_save(insetmap, filename="../03_Outputs/inset.svg",
          dpi=600, scale = 1, width=5, units="cm")
```


# Fig extra: Raincloud plots {.tabset}

To visualise the environmental data in the presence and absence of sperm whales, raincloud plots are used.  
Don't want those to be on the standardised scale, so I start from the original dataset, which I strip down to keep only the variables retained in the model (`slope2.5`, `aspect2.5`, `tpi10`, `distfront7km`, `weeksst4km`, `SLA`) and the `PA` column.

```{r}
datavis <- readRDS("../01_Data/SPW_PAM_allcovariates.rds")
datavis <- cbind(st_coordinates(datavis), as.data.frame(st_drop_geometry(datavis)))
select <- c("PA", "slope2.5", "aspect2.5", "tpi10", "distfront7km", "weeksst4km", "SLA")
datavis <- datavis[,select]
```

The individual plots for each variable are found in the tabs below.

## Slope {-}
```{r}
theme_set(theme_bw(16)) # apparently need that for the half violins
source("https://raw.githubusercontent.com/datavizpyr/data/master/half_flat_violinplot.R")

scaleFUN <- function(x) sprintf("%.1f", x) # custom function to keep trailing zeros in axis labels
slope <- ggplot(data = datavis, aes(x = as.factor(PA), 
                                      y = slope2.5, 
                                      fill = as.factor(PA),
                                      colour = as.factor(PA))) +
  geom_flat_violin(alpha = .9, position = position_nudge(x = .12, y = 0)) +
 # stat_ydensity(geom = "bar", colour = "white", fill = "white", width = .3) +
  geom_jitter(width = .1, 
              size = .3, 
              alpha = 0.25,
              fill = NA,
              col = "grey30") +
  geom_boxplot(width = 0.1, 
               outlier.shape = NA, 
               alpha = 0.6,
               color = "black") +
  labs(y="Slope (°)", x=" ") +
  ggtitle("a") +
  coord_flip() +
  theme_classic() +
  theme(legend.position = "none", axis.title.x = element_text(size=13, family = 'serif'),
        axis.text.y = element_text(angle=90, hjust=-0.05, size=13),
        axis.text=element_text(size=11, family = 'serif'), axis.ticks.y = element_blank(),
        plot.title = element_text(family = "serif", size = 13, vjust = -5, hjust = 0.02),
        plot.margin = margin(t=30, b=5, r=1)) +
  scale_x_discrete(expand = c(0.15, 0), labels=c("Absence", "Presence")) +
  scale_y_continuous(limits = c(0, 13), breaks = c(0,4,8,12)) +
  scale_colour_manual(values = c("#dae5eb", "#4E6D82")) +
  scale_fill_manual(values = c("#dae5eb", "#4E6D82"))

print(slope)
```

Here are some summary statistics for `slope2.5` values when sperm whales were absent:
```{r}
summary(datavis$slope2.5[which(datavis$PA == 0)])
```

And when sperm whales were present:
```{r}
summary(datavis$slope2.5[which(datavis$PA == 1)])
```

## Aspect {-}
```{r, echo=FALSE}
aspect <- ggplot(data = datavis, aes(x = as.factor(PA), 
                                      y = aspect2.5, 
                                      fill = as.factor(PA),
                                      colour = as.factor(PA))) +
  geom_flat_violin(alpha = .9, position = position_nudge(x = .12, y = 0)) +
#  stat_ydensity(geom = "bar", colour = "white", fill = "white", width = .3) +
  geom_jitter(width = .1, 
              size = .3, 
              alpha = 0.25,
              fill = NA,
              col = "grey30") +
  geom_boxplot(width = 0.1, 
               outlier.shape = NA, 
               alpha = 0.6,
               color = "black") +
  labs(y="Aspect (°)", x="") +
  ggtitle("b") +
  coord_flip() +
  theme_classic() +
  theme(legend.position = "none", axis.title.x = element_text(size=13, family = 'serif'),
        axis.text.y = element_text(angle=90, hjust=-0.05),
        axis.text=element_text(size=11, family = 'serif'), axis.ticks.y = element_blank(),
        plot.title = element_text(family = "serif", size = 13, vjust = -5, hjust = 0.02),
        plot.margin = margin(t=30, b=5, r=1)) +
  scale_x_discrete(expand = c(0.15, 0), labels=c("", "")) +
  scale_y_continuous(breaks=c(0, 120, 240, 360)) +
  scale_colour_manual(values = c("#dae5eb", "#BFDFED")) +
  scale_fill_manual(values = c("#dae5eb", "#BFDFED"))

print(aspect)
```

Here are some summary statistics for `aspect2.5` values when sperm whales were absent:
```{r}
summary(datavis$aspect2.5[which(datavis$PA == 0)])
```

And when sperm whales were present:
```{r}
summary(datavis$aspect2.5[which(datavis$PA == 1)])
```

## TPI {-}
```{r echo=FALSE, results = TRUE}
tpi <- ggplot(data = datavis, aes(x = as.factor(PA), 
                                      y = tpi10, 
                                      fill = as.factor(PA),
                                      colour = as.factor(PA))) +
  geom_flat_violin(alpha = .9, position = position_nudge(x = .12, y = 0)) +
#  stat_ydensity(geom = "bar", colour = "white", fill = "white", width = .3) +
  geom_jitter(width = .10, 
              size = .3, 
              alpha = 0.25,
              fill = NA,
              col = "grey30") +
  geom_boxplot(width = 0.1, 
               outlier.shape = NA, 
               alpha = 0.6,
               color = "black") +
  labs(y="Topographic position index (m)", x="") +
  ggtitle("c") +
  coord_flip() +
  theme_classic() +
  theme(legend.position = "none", axis.title.x = element_text(size=13, family = 'serif'),
        axis.text.y = element_text(angle=90, hjust=-0.05),
        axis.text=element_text(size=11, family = 'serif'), axis.ticks.y = element_blank(),
        plot.title = element_text(family = "serif", size = 13, vjust = -5, hjust = 0.02),
        plot.margin = margin(t=30, b=5, r=1)) +
  scale_x_discrete(expand = c(0.15, 0), labels=c("", "")) +
  scale_colour_manual(values = c("#dae5eb", "#5DA5AC")) +
  scale_fill_manual(values = c("#dae5eb", "#5DA5AC"))

print(tpi)
```

Here are some summary statistics for `tpi10` values when sperm whales were absent:
```{r}
summary(datavis$tpi10[which(datavis$PA == 0)])
```

And when sperm whales were present:
```{r}
summary(datavis$tpi10[which(datavis$PA == 1)])
```

## Distant to front {-}
```{r, echo=FALSE}
distfront <- ggplot(data = datavis, aes(x = as.factor(PA), 
                                 y = distfront7km, 
                                 fill = as.factor(PA),
                                 colour = as.factor(PA))) +
  geom_flat_violin(alpha = .9, position = position_nudge(x = .12, y = 0)) +
#  stat_ydensity(geom = "bar", colour = "white", fill = "white", width = .3) +
  geom_jitter(width = .10, 
              size = .3, 
              alpha = 0.25,
              fill = NA,
              col = "grey30") +
  geom_boxplot(width = 0.1, 
               outlier.shape = NA, 
               alpha = 0.6,
               color = "black") +
  labs(y="Distance to front (km)", x="") +
  ggtitle("e") +
  coord_flip() +
  theme_classic() +
  theme(legend.position = "none", axis.title.x = element_text(size=13, family = 'serif'),
        axis.text.y = element_text(angle=90, hjust=-0.04),
        axis.text=element_text(size=11, family = 'serif'), axis.ticks.y = element_blank(),
        plot.title = element_text(family = "serif", size = 13, vjust = -5, hjust = 0.02),
        plot.margin = margin(t=30, b=5, r=1)) +
  scale_x_discrete(expand = c(0.15, 0), labels=c("", "")) +
#  scale_y_continuous(limit = range(data$depth2.5), breaks = c(0.25,0.75,1.25)) +
  scale_colour_manual(values = c("#dae5eb", "#EBC0AD")) +
  scale_fill_manual(values = c("#dae5eb", "#EBC0AD"))

print(distfront)
```

Here are some summary statistics for `distfront7km` values when sperm whales were absent:
```{r}
summary(datavis$distfront7km[which(datavis$PA == 0)])
```

And when sperm whales were present:
```{r}
summary(datavis$distfront7km[which(datavis$PA == 1)])
```

## SST {-}
```{r, echo=FALSE}
sst <- ggplot(data = datavis, aes(x = as.factor(PA), 
                                 y = weeksst4km, 
                                 fill = as.factor(PA),
                                 colour = as.factor(PA))) +
  geom_flat_violin(alpha = .9, position = position_nudge(x = .12, y = 0)) +
#  stat_ydensity(geom = "bar", colour = "white", fill = "white", width = .3) +
  geom_jitter(width = .10, 
              size = .3, 
              alpha = 0.25,
              fill = NA,
              col = "grey30") +
  geom_boxplot(width = 0.1, 
               outlier.shape = NA, 
               alpha = 0.6,
               color = "black") +
  labs(y="Sea surface temperature (°C)", x="") +
  ggtitle("d") +
  coord_flip() +
  theme_classic() +
  theme(legend.position = "none", axis.title.x = element_text(size=13, family = 'serif'),
        axis.text.y = element_text(angle=90, hjust=-0.04, size=13),
        axis.text=element_text(size=11, family = 'serif'), axis.ticks.y = element_blank(),
        plot.title = element_text(family = "serif", size = 13, vjust = -5, hjust = 0.02),
        plot.margin = margin(t=30, b=5, r=1)) +
  scale_x_discrete(expand = c(0.15, 0), labels=c("Absence", "Presence")) +
#  scale_y_continuous(limit = range(data$depth2.5), breaks = c(0.25,0.75,1.25)) +
  scale_colour_manual(values = c("#dae5eb", "#F4CE92")) +
  scale_fill_manual(values = c("#dae5eb", "#F4CE92"))

print(sst)
```

Here are some summary statistics for `weeksst4km` values when sperm whales were absent:
```{r}
summary(datavis$weeksst4km[which(datavis$PA == 0)])
```

And when sperm whales were present:
```{r}
summary(datavis$weeksst4km[which(datavis$PA == 1)])
```


## SLA {-}
```{r, echo=FALSE}
scaleFUN <- function(x) sprintf("%.2f", x) # custom function to keep trailing zeros in axis labels
SLA <- ggplot(data = datavis, aes(x = as.factor(PA), 
                                 y = SLA, 
                                 fill = as.factor(PA),
                                 colour = as.factor(PA))) +
  geom_flat_violin(alpha = .9, position = position_nudge(x = .12, y = 0)) +
#  stat_ydensity(geom = "bar", colour = "white", fill = "white", width = .3) +
  geom_jitter(width = .10, 
              size = .3, 
              alpha = 0.25,
              fill = NA,
              col = "grey30") +
  geom_boxplot(width = 0.1, 
               outlier.shape = NA, 
               alpha = 0.6,
               color = "black") +
  labs(y="Sea level anomaly (m)", x="") +
  ggtitle("f") +
  coord_flip() +
  theme_classic() +
  theme(legend.position = "none", axis.title.x = element_text(size=13, family = 'serif'),
        axis.text.y = element_text(angle=90, hjust=-0.04),
        axis.text=element_text(size=11, family = 'serif'), axis.ticks.y = element_blank(),
        plot.title = element_text(family = "serif", size = 13, vjust = -5, hjust = 0.02),
        plot.margin = margin(t=30, b=5, r=1)) +
  scale_x_discrete(expand = c(0.15, 0), labels=c("", "")) +
  scale_y_continuous(breaks = c(-0.15, -0.1, -0.05, 0, 0.05, 0.10), labels = scaleFUN) +
  scale_colour_manual(values = c("#dae5eb", "#957580")) +
  scale_fill_manual(values = c("#dae5eb", "#957580"))

print(SLA)
```

Here are some summary statistics for `SLA` values when sperm whales were absent:
```{r}
summary(datavis$SLA[which(datavis$PA == 0)])
```

And when sperm whales were present:
```{r}
summary(datavis$SLA[which(datavis$PA == 1)])
```

# {-}

```{r echo=F, results='hide', fig.keep='all', out.width="80%"}
# To save them altogether in one figure directly
#cairo_ps(filename = "../03_Outputs/Fig.3_Observed.eps", fallback_resolution = 600, width = 10, height = 6) # Fallback_resolution for the transparency parts
cowplot::plot_grid(slope, aspect, tpi, sst, distfront, SLA, nrow = 2, align = "hv")
#dev.off()
```

# Fig. 4: Partial plots

Initially, those below are run in the `03_Modelling.Rmd`. I copied it here too to have all figures in one place.

```{r, results='hide', fig.keep='all', out.width="90%"}
data <- readRDS("../01_Data/data_model.rds")
pa_col <- 7 # presence absence column
var_cols <- c(9, 11, 13, 10, 16, 14, 8, 15, 12) # change the order
names(data[var_cols]) # double check

mod_bart_final <- bart(y.train = data[ , pa_col],
                  x.train = data[, var_cols],
                  keeptrees = TRUE, verbose = FALSE)

source("Partialplots_fun.R") # load functions

aspect.pp <- aspect.partial(mod_bart, x.vars = "aspect2.5", ciwidth = 0.95, 
                            trace=FALSE)
front.pp <- distfront.partial(mod_bart, x.vars = "distfront7km", ciwidth = 0.95,
                              trace=FALSE)
sla.pp <- sla.partial(mod_bart, x.vars = "SLA", ciwidth = 0.95, 
                      trace=FALSE)
tpi.pp <- tpi.partial(mod_bart, x.vars = "tpi10", ciwidth = 0.95, 
                      trace=FALSE)
relsst.pp <- relsst.partial(mod_bart, x.vars = "relsst.month4km", ciwidth = 0.95, 
                            trace=FALSE)
slope.pp <- slope.partial(mod_bart, x.vars = "slope2.5", ciwidth = 0.95, 
                          trace=FALSE)
depth.pp <- depth.partial(mod_bart, x.vars = "depth2.5", ciwidth = 0.95, 
                          trace=FALSE)
chla.pp <- chla.partial(mod_bart, x.vars = "chla4km", ciwidth = 0.95, 
                        trace=FALSE)

# cairo_ps(filename = "../03_Outputs/Fig.4_PartialPlots.eps", fallback_resolution = 600, width = 13, height = 6)
cowplot::plot_grid(aspect.pp[[1]], front.pp[[1]], sla.pp[[1]], tpi.pp[[1]], 
                   relsst.pp[[1]], slope.pp[[1]], depth.pp[[1]], chla.pp[[1]],
                   nrow = 2, align = "hv")
# dev.off()
```

# Fig. 4: Favourability maps

```{r load data}
favourability.raster <- 
  rast("F:/PhD/Ch4_Modelling_SPW_PAM/01_Data/Rasters/final_pred_favourability.tif")
```

Let's start with the base map that will be common to all favourability maps: `base`.

```{r basemap, results='hide', fig.keep = 'all'}
# define bounding box for the plot area
area.box <- as(extent(-16, -8, 48.5, 56.5), 'SpatialPolygons') # create a box as a Spatial Object
crs(area.box) <- "+proj=longlat +datum=WGS84 +no_defs" # define the crs of this box
box <- spTransform(area.box, CRS("+init=epsg:32629")) # project the spatial object (WGS84 UTM Z29: EPSG 32629)

bbox_new <- st_bbox(box) %>%  # take the bounding box ...
  st_as_sfc() # ... and make it a sf polygon

base <- gisco_get_countries(country = c("Ireland", "United Kingdom"), 
                               year = "2020", epsg = "4326",
                               resolution = 1, spatialtype = "RG") %>%
  st_transform(projection) # project it

base <- tm_shape(base, bbox = bbox_new) +
  tm_borders(lwd = 0.5, col = "grey10") +
  tm_fill(col="grey20")

print(base)
```

Now let's add the elements.

```{r elements, results='hide', fig.keep = 'all'}
graticules <- tm_graticules(lines = FALSE)

# element with scale only for the last plot, as to not repeat it all the time
scale <- tm_scale_bar(position=c("RIGHT", "BOTTOM"), breaks = c(0,100,200), lwd = 0.1)

print(base + graticules + scale)
```

And now we need to create the layers which will be overlaid, corresponding to the predictions in the different months/seasons of interest: `spring`, `june`, `august` and `autumn`. 

```{r prediction layers, results='hide', fig.keep = 'all'}
# palette
library("colorspace")
library("viridis")
pal1 <- rev(magma(100, alpha = 0.8))
pal <- rev(sequential_hcl(100, palette = "Purple-Yellow"))

# use this one instead
brks <- seq(0,1,0.1)

spring <- tm_shape(favourability.raster$spring) +
  tm_raster(palette = my_palette(100), style = "cont", breaks = brks, legend.show = FALSE) +
  tm_credits("a", fontfamily = "serif", position = c("LEFT", "BOTTOM"), size = 0.9)

june <- tm_shape(favourability.raster$june) +
  tm_raster(palette = my_palette(100), style = "cont", breaks = brks, legend.show = FALSE)+
  tm_credits("b", fontfamily = "serif", position = c("LEFT", "BOTTOM"), size = 0.9)

august <- tm_shape(favourability.raster$august) +
  tm_raster(palette = my_palette(100), style = "cont", breaks = brks, legend.show = FALSE) +
  tm_credits("c", fontfamily = "serif", position = c("LEFT", "BOTTOM"), size = 0.9)

autumn <- tm_shape(favourability.raster$autumn) +
  tm_raster(palette = my_palette(100), style = "cont", breaks = brks, legend.show = F) +
  tm_credits("d", fontfamily = "serif", position = c("LEFT", "BOTTOM"), size = 0.9)

print(base + graticules + scale + autumn)
```

To make it look a bit nicer I want to add a box that delimits the predictions zone. I did that in the Modelling file (`03_Modelling.Rmd`) so I'll repeat here. The tmap elements is called `zone`.

```{r predictions zone, results='hide', fig.keep = 'all', warning=FALSE, message=FALSE}
# get the bounding boxes
northbox <- st_read("F:/PhD/ObSERVE_MASTER/ObSERVE_PAM/Shapefiles/NorthBox.shp")
southbox <- st_read("F:/PhD/ObSERVE_MASTER/ObSERVE_PAM/Shapefiles/SouthBox.shp")
# Merge them
area <- st_union(northbox, southbox)

# Create a polygon encompassing both zones, with a little buffer around
pred.zone <- st_simplify(st_buffer(area, dist = 20000), dTolerance = 7000)
# extended slightly the buffer so it goes well around the limits of the predictions
zone <- 
  tm_shape(pred.zone) +
    tm_borders(col="grey20", lwd = 1.5, lty = 1)

print(base + graticules + scale + spring + zone)
```

That looks quite good. But the transect lines would be adding an extra piece of information there, so let's add them. Only the tracks corresponding to the season/month in which the predictions were computed. Four different map elements will be created: `track.spring`, `track.june`, `track.august` and `track.autumn`.

```{r add transects, results='hide', fig.keep = 'all', warning=FALSE, message=FALSE}
track.spring <- transects[which(transects$Survey == 1 | transects$Survey == 4) ,]
track.june <- transects[which(transects$Survey == 5) ,]
track.august <- transects[which(transects$Survey == 2) ,]
track.autumn <- transects[which(transects$Survey == 3 | transects$Survey == 6) ,]

track.spring <-
  tm_shape(track.spring) +
    tm_lines(col="grey15", lwd = 0.5, legend.col.show = F)

track.june <-
  tm_shape(track.june) +
    tm_lines(col="grey15", lwd = 0.5, legend.col.show = F)

track.august <-
  tm_shape(track.august) +
    tm_lines(col="grey15", lwd = 0.5, legend.col.show = F)

track.autumn <-
  tm_shape(track.autumn) +
    tm_lines(col="grey15", lwd = 0.5, legend.col.show = F)

print(track.spring)
```

The sightings as well:

```{r}
det.spring <- det[which(det$month_ == 5) , ]
det.june <- det[which(det$month_ == 6) , ]
det.august <- det[which(det$month_ == 8) , ]
det.autumn <- det[which(det$month_ == 9 | det$month_ == 10) , ]

det.spring <- 
  tm_shape(det.spring) +
    tm_symbols(size = 0.1, col = "grey15", border.col = "grey15", alpha = 0.4, 
               border.lwd = 0.1, legend.col.show = F)
det.june <- 
    tm_shape(det.june) +
    tm_symbols(size = 0.1, col = "grey15", border.col = "grey15", alpha = 0.4, 
               border.lwd = 0.1, legend.col.show = F)
det.august <- 
    tm_shape(det.august) +
    tm_symbols(size = 0.1, col = "grey15", border.col = "grey15", alpha = 0.4, 
               border.lwd = 0.1, legend.col.show = F)
det.autumn <- 
    tm_shape(det.autumn) +
    tm_symbols(size = 0.1, col = "grey15", border.col = "grey15", alpha = 0.4, 
               border.lwd = 0.1, legend.col.show = F)

print(det.spring)
```

Let's also add some titles to the maps:

```{r map titles, results='hide', fig.keep = 'all', warning=FALSE, message=FALSE}
title.spring <- tm_layout(title = " May 2015-2016", title.size = 0.9,
                          title.bg.color = "white", title.position = c("LEFT", "TOP"),
                          title.fontfamily = "serif")
title.june <- tm_layout(title = " June 2016", title.size = 0.9,
                          title.bg.color = "white", title.position = c("LEFT", "TOP"),
                          title.fontfamily = "serif")
title.august <- tm_layout(title = " August 2015", title.size = 0.9,
                          title.bg.color = "white", title.position = c("LEFT", "TOP"),
                          title.fontfamily = "serif")
title.autumn <- tm_layout(title = " Sept-Oct 2015-2016", title.size = 0.9,
                          title.bg.color = "white", title.position = c("LEFT", "TOP"),
                          title.fontfamily = "serif")
```

For the legend, I want the terra version of it. I didn't find a work around so I'm exporting the image below from the plot window and will cut and add the legend myself to my plot!

```{r legend preds, results='hide'}
legend <- terra::plot(favourability.raster$autumn, col=pal, cex=16)
```

Now let's create the final maps. Only one of them is displayed.

```{r final maps, results='hide'}
fav.spring <- base + graticules + spring + zone + track.spring + det.spring + 
  title.spring
fav.june <- base + graticules + june + zone + track.june + det.june + title.june
fav.august <- base + graticules + august + zone + track.august + det.august + 
  title.august
fav.autumn <- base + graticules + scale + autumn + zone + track.autumn + 
  det.autumn + title.autumn

print(fav.autumn)
```

Save all elements. I'm saving them all as `.png` and not `.svg` because the latter generates files that are way too heavy (~88 MB).

```{r eval=FALSE}
# print bbox_new to get the limits to do: xmax - xmin / ymax - ymin
asp <- (573871.4 - (-16944.78))/(6283704 - 5372358)
tmap_save(pred.spring, filename="../03_Outputs/fig5.spring.png",
          dpi=900, height=asp*15, width=15, units="cm")
tmap_save(pred.june, filename="../03_Outputs/fig5.june.png",
          dpi=900, height=asp*15, width=15, units="cm")
tmap_save(pred.august, filename="../03_Outputs/fig5.aug.png",
          dpi=900, height=asp*15, width=15, units="cm")
tmap_save(pred.autumn, filename="../03_Outputs/fig5.aut.png",
          dpi=900, height=asp*15, width=15, units="cm")
```

# Extra: Predictions and uncertainty

```{r prep}
predictions.rast <- rast("F:/PhD/Ch4_Modelling_SPW_PAM/01_Data/Rasters/final_predictions.tif")
my_palette <- colorRampPalette(c("#264653", "#86c9dd", "#e5b761", "#efd199"))
```

We can re-use all the elements from before and just add the new raster layers to replace that of the favourability.
So let's just prepare the prediction layers:

```{r pred uncert layers}
spring.pred <- tm_shape(predictions.rast$spring) +
  tm_raster(palette = my_palette(100), style = "cont", breaks = brks, legend.show = FALSE) +
  tm_credits("a", fontfamily = "serif", position = c("LEFT", "BOTTOM"), size = 0.9)

june.pred <- tm_shape(predictions.rast$june) +
  tm_raster(palette = my_palette(100), style = "cont", breaks = brks, legend.show = FALSE)+
  tm_credits("b", fontfamily = "serif", position = c("LEFT", "BOTTOM"), size = 0.9)

august.pred <- tm_shape(predictions.rast$august) +
  tm_raster(palette = my_palette(100), style = "cont", breaks = brks, legend.show = FALSE) +
  tm_credits("c", fontfamily = "serif", position = c("LEFT", "BOTTOM"), size = 0.9)

autumn.pred <- tm_shape(predictions.rast$autumn) +
  tm_raster(palette = my_palette(100), style = "cont", breaks = brks, legend.show = F) +
  tm_credits("d", fontfamily = "serif", position = c("LEFT", "BOTTOM"), size = 0.9)


# Compute uncertainty layers
spring.uncert <- tm_shape(predictions.rast$spring.q2 - predictions.rast$spring.q1) +
  tm_raster(palette = my_palette(100), style = "cont", breaks = brks, legend.show = FALSE) +
  tm_credits("a", fontfamily = "serif", position = c("LEFT", "BOTTOM"), size = 0.9)

june.pred <-  tm_shape(predictions.rast$june.q2 - predictions.rast$june.q1) +
  tm_raster(palette = my_palette(100), style = "cont", breaks = brks, legend.show = FALSE) +
  tm_credits("a", fontfamily = "serif", position = c("LEFT", "BOTTOM"), size = 0.9)

august.pred <-  tm_shape(predictions.rast$august.q2 - predictions.rast$august.q1) +
  tm_raster(palette = my_palette(100), style = "cont", breaks = brks, legend.show = FALSE) +
  tm_credits("a", fontfamily = "serif", position = c("LEFT", "BOTTOM"), size = 0.9)

autumn.pred <-  tm_shape(predictions.rast$autumn.q2 - predictions.rast$autumn.q1) +
  tm_raster(palette = my_palette(100), style = "cont", breaks = brks, legend.show = FALSE) +
  tm_credits("a", fontfamily = "serif", position = c("LEFT", "BOTTOM"), size = 0.9)
```



```{r pred maps, results='hide'}
pred.spring <- base + graticules + spring + zone + track.spring + det.spring + title.spring
```